Learning Libraries of Subroutines for Neurally--Guided Bayesian Program Learning

### 2018-05-16 nips_2018.tex
better introduction figures

Our algorithm is called \textbf{Explore/Compress/Compile} (\systemEnding),
because it iterates between three different steps: an
\textbf{Explore} step uses the DSL to explore the space of programs,
searching for ones that solve the tasks; a \textbf{Compress} step
modifies the DSL by discovering regularities in the programs found by
the previous Explore step; and a \textbf{Compile} step, which
improves the program search procedure by training a neural network to
write programs in the current DSL, in the spirit of ``amortized'' or
``compiled'' inference~\cite{le2016inference}.
We call the neural net a \textbf{recognition model} (c.f. Hinton 1995~\cite{hinton1995wake}).
The learned DSL
distills commonalities across programs that solve tasks, helping
the agent solve related program induction problems. The neural
recognition model ensures that searching for programs remains tractable
even as the DSL (and hence the search space for programs) expands.

Because any model may be encoded as a (deterministic or probabilistic) program,
we carefully delineate the scope of program learning problems considered here.
We think of \system as learning to solve the kinds of problems that humans
can solve relatively quickly -- once they acquire the relevant domain expertise.
These correspond to short programs if you have an expressive DSL.
Even with such an expressive DSL, program search may be intractable, so
we amortize the cost of program search by training a neural network to
assist the search procedure.

We apply \system to four domains:
list processing; FlashFill-style~\cite{gulwani2011automating} string editing;  symbolic regression; and turtle graphics~\cite{turtle}.
 For each of these we initially provide a generic
 set of programming primitives.
Our algorithm then discovers
its own domain-specific vocabulary for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).
\begin{wrapfigure}{r}{0.6\textwidth}
    \begin{tabular}{ll}
   \toprule
   Domain&Part of the learned DSL\\\midrule
   Lists& \code{(foldr nil (lambda (a b) (cons a b)))}\\
   &\hspace{0.5cm} \emph{(reverses a list)}\\
   Strings& \code{(map (lambda (x) (if (= x a) b x)))}\\
   &\hspace{0.5cm} \emph{(replace occurrences of a w/ b)}\\
   Regression& \code{(+ (* real x) real)} \\
      &\hspace{0.5cm} \emph{(a linear function of x)}\\
   \\\bottomrule
    \end{tabular}
    \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful subroutines.}\label{initialExampleDSL}
    \vspace{-4em}
 \end{wrapfigure}

@@

 Our goal is to induce a DSL while finding programs solving each of the
tasks.  We take inspiration primarily from the
Exploration-Compression algorithm for bootstrap
learning~\cite{Dechter:2013:BLV:2540128.2540316}.  Exploration-Compression
alternates between exploring the space of solutions to a set of tasks,
and compressing those solutions to suggest new search primitives
for the next exploration stage.
We extend these ideas into an inference strategy that iterates
through three steps: an \textbf{Explore} step uses the current DSL and
recognition model to search for programs that solve the tasks.  The
\textbf{Compress} and \textbf{Compile} steps update the DSL and the
recognition model, respectively.  Crucially, these steps bootstrap 
off each other (Fig.~\ref{feeding}):

\begin{wrapfigure}{r}{0.5\textwidth}
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs};
    \node[align = center](DSL) at (3,1) {DSL};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model};

@@

    \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Inductive bias}\\\footnotesize{(Exploration)}} (synthesis.west);
    \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white,align = center]{{\footnotesize Makes tractable}\\{\footnotesize (Exploration)}} (synthesis.east);
  \end{scope}
    \end{tikzpicture}
  \caption{\system solves for programs, the DSL, and a neural network (recognition model). Each of these steps iteratively bootstraps off of the others.}  \label{feeding}\vspace{-0.4cm}
\end{wrapfigure}%
\\\noindent \textbf{Exploration: Searching for programs.}  Our program search
is informed by both the DSL and the recognition model. When these
improve, we find more programs solving the tasks.
\\\noindent\textbf{Compression: Improving the DSL.} We induce the DSL
from the programs found in the exploration phase, aiming
to maximally compress (or, raise the prior probability of) these
programs.  As we solve more tasks, we hone in on DSLs that
more closely match the domain.
\\\noindent \textbf{Compilation: Learning a neural recognition model.}  We
update the recognition model by training on two data sources: samples
from the DSL (as in the Helmholtz Machine's ``sleep'' phase), and
programs found by the search procedure during exploration. As the DSL
improves and as search finds more programs, the recognition model gets
more data to train on, and better data.

@@

\system takes as input a set of \emph{tasks}, written $X$, each of which is a program induction problem.
It has at its disposal a \emph{likelihood model}, written $\probability[x|p]$, which scores the likelihood of a task $x\in X$ given a program $p$.
Its goal is to solve each of the tasks by writing a program,
and also to infer a DSL, written $\mathcal{D}$.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a generative model over programs.
We frame our goal as maximum a posteriori (MAP) inference of $(\mathcal{D},\theta)$ given $X$.
Writing $J$ for the joint probability of $(\mathcal{D},\theta)$ and $X$, we want the $\mathcal{D}^*$ and $\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)\triangleq \probability[\mathcal{D},\theta]\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* = \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta \qquad
  \theta^* =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}

>>>>

Because any model may be encoded as a (deterministic or probabilistic) program,
we carefully delineate the scope of program learning problems considered here.
We think of \system as learning to solve the kinds of problems that humans
can solve relatively quickly -- once they acquire the relevant domain expertise.
These correspond to short programs -- if you have an expressive DSL.
Even with a good DSL, program search may be intractable, so
we amortize the cost of program search by training a neural network to
assist the search procedure.

Our algorithm is called \textbf{Explore/Compress/Compile} (\systemEnding),
because it iterates between three different steps: an
\textbf{Explore} step uses the DSL to explore the space of programs,
searching for ones that solve the tasks; a \textbf{Compress} step
modifies the structure of the DSL by discovering regularities across programs found by
the previous Explore step; and a \textbf{Compile} step, which
improves the program search procedure by training a neural network to
write programs in the current DSL, in the spirit of ``amortized'' or
``compiled'' inference~\cite{le2016inference}.
We call the neural net a \textbf{recognition model} (c.f. Hinton 1995~\cite{hinton1995wake}).
The learned DSL
distills commonalities across programs that solve tasks, helping
the agent solve related program induction problems. The neural
recognition model ensures that searching for programs remains tractable
even as the DSL (and hence the search space for programs) expands.

We apply \system to four domains:
list processing; FlashFill-style~\cite{gulwani2011automating} string editing;  symbolic regression; and turtle graphics~\cite{turtle}.
 For each of these we initially provide a generic
 set of programming primitives.
Our algorithm then discovers
its own domain-specific vocabulary for expressing solutions in the domain (Tbl.~\ref{initialExampleDSL}).
\begin{figure}
  \begin{tabular}{lcl}
    \toprule
    Domain&Example Task&Part of the learned DSL\\\midrule
    Lists&
    \begin{tabular}{rcl}
      \code{[7\, 0\, 2]}&$\longrightarrow$&\code{[7\, 0\, 2\, 4]} \\
      \code{[3\, 9]}&$\longrightarrow$&\code{[3\, 9\, 4]}
    \end{tabular}
    &
    \begin{tabular}{l}
      \code{(foldr nil (lambda (a b) (cons a b)))}\\
          \hspace{0.5cm} \emph{(appends lists)}
      \end{tabular}

    \\    Strings&
    \begin{tabular}{rcl}
      Temple Anna H& $\longrightarrow$& TAH\\
      Lara Gregori&$\longrightarrow$&LG
      \end{tabular}
    &
    \begin{tabular}{l}
      \code{(map (lambda (x) (if (= x a) b x)))}\\
      \hspace{0.5cm} \emph{(replace occurrences of a w/ b)}\\
      \end{tabular}

    \\Regression&\includegraphics[width = 1cm]{figures/functions/476.png}&\begin{tabular}{l}
      \code{(+ (* real x) real)} \\
          \hspace{0.5cm} \emph{(a linear function of x)}
      \end{tabular}

    \\\bottomrule
  \end{tabular}
  \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful subroutines.}\label{initialExampleDSL}
%  \vspace{-4em}
\end{figure}

@@

 Our goal is to induce a DSL while finding programs solving each of the
tasks.  We take inspiration primarily from the
Exploration-Compression algorithm for bootstrap
learning~\cite{Dechter:2013:BLV:2540128.2540316}.  Exploration-Compression
alternates between exploring the space of solutions to a set of tasks,
and compressing those solutions to suggest new search primitives
for the next exploration stage.
We extend these ideas into an inference strategy that iterates
through three steps: an \textbf{Explore} step uses the current DSL and
recognition model to search for programs that solve the tasks.  The
\textbf{Compress} and \textbf{Compile} steps update the DSL and the
recognition model, respectively.  Crucially, these steps synergistically bootstrap 
off each other:
\begin{comment}
  \begin{wrapfigure}{r}{0.5\textwidth}
  \begin{tikzpicture}
    \begin{scope}[shift = {(1,-1)}]
    \node[align = center](synthesis) at (6,4) {Search for \\programs};
    \node[align = center](DSL) at (3,1) {DSL};
    \node[align = center](recognitionModel) at (9,1) {Recognition \\model};

@@

    \draw [->,thick,dashed] (DSL.north) to[out = 90,in = 180] node[fill=white,align = center]{  \footnotesize{Inductive bias}\\\footnotesize{(Exploration)}} (synthesis.west);
    \draw [->,thick,dashed] (recognitionModel.north) to[out = 90,in = 0] node[fill=white,align = center]{{\footnotesize Makes tractable}\\{\footnotesize (Exploration)}} (synthesis.east);
  \end{scope}
    \end{tikzpicture}
  \caption{\system solves for programs, the DSL, and a neural network (recognition model). Each of these steps iteratively bootstraps off of the others.}  \label{feeding}\vspace{-0.4cm}
\end{wrapfigure}
  \end{comment}
\\\noindent \textbf{Exploration: Searching for programs.}  Our program search
is informed by both the DSL and the recognition model. When these
improve, we can solve more tasks.
\\\noindent\textbf{Compression: Improving the DSL.} We induce the DSL
from the programs found in the exploration phase, aiming
to maximally compress (or, raise the prior probability of) these
programs.  As we solve more tasks, we hone in on DSLs that
more closely match the domain.
\\\noindent \textbf{Compilation: Learning a neural recognition model.}  We
update the recognition model by training on two data sources: samples
from the DSL (as in the Helmholtz Machine's ``sleep'' phase), and
programs found by the search procedure during exploration. As the DSL
improves and as search finds more programs, the recognition model gets
more data to train on, and better data.

@@

\system takes as input a set of \emph{tasks}, written $X$, each of which is a program synthesis problem.
It has at its disposal a domain-specific \emph{likelihood model}, written $\probability[x|p]$, which scores the likelihood of a task $x\in X$ given a program $p$.\footnote{For example, for string editing,
  the likelihood is 1 if the program predicts the observed outputs on the observed inputs,
and 0 otherwise.}
Its goal is to solve each of the tasks by writing a program,
and also to infer a DSL, written $\mathcal{D}$.
We equip $\mathcal{D}$ with a real-valued weight vector $\theta$, and together
$(\mathcal{D},\theta)$ define a generative model over programs.
We frame our goal as maximum a posteriori (MAP) inference of $(\mathcal{D},\theta)$ given $X$.
Writing $J$ for the joint probability of $(\mathcal{D},\theta)$ and $X$, we want the $\mathcal{D}^*$ and $\theta^*$ solving:
\begin{align}\label{intractableObjectives}
\nonumber  J(\mathcal{D},\theta)\triangleq \probability[\mathcal{D},\theta]\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\\
  \mathcal{D}^* = \argmax_{\mathcal{D}}\int J(\mathcal{D},\theta)\;\mathrm{d}\theta \qquad
  \theta^* =\argmax_\theta J(\mathcal{D}^*,\theta)
\end{align}
